---
title: "Healthcare_cost_analysis"
author: 'BY: KRISHNA ARYAL'
date: "8/8/2020"
output:
  pdf_document: default
  classoption: a4paper
  documentclass: article
  fontsize: 36
---
\fontsize{12}{22}
\selectfont

# $\color{blue}{\text{Healthcare cost analysis}}$
$\color{red}{\text{Background and Objective:}}$
A nationwide survey of hospital costs conducted by the US Agency for Healthcare consists of hospital records of inpatient samples. The given data is restricted to the city of Wisconsin and relates to patients in the age group 0-17 years. The agency wants to analyze the data to research on healthcare costs and their utilization.

Domain: Healthcare

Dataset Description:

Here is a detailed description of the given dataset:

Attribute	Description
1. Age 	Age of the patient discharged

2. Female 	A binary variable that indicates if the patient is female

3. Los	Length of stay in days

4. Race 	
Race of the patient (specified numerically)

5. Totchg	Hospital discharge costs

6. Aprdrg	All Patient Refined Diagnosis Related Groups

\newpage
$\color{red}{\text{ANALYSIS TO BE DONE:}}$ 

### 1. TO RECORD THE PATIENT STATISTICS, THE AGENCY WANTS TO FIND THE AGE CATEGORY OF PEOPLE WHO FREQUENT THE HOSPITAL AND HAS THE MAXIMUM EXPENDITURE.
```{r}
hos_cost <- read.csv("HospitalCosts.csv")
#TO FIND AGE CATEGORY OF PEOPLE WHO VISITED HOSPITAL FREQUENTLY
Age<-hos_cost$AGE
hist(Age,freq = TRUE,right = TRUE,border = "red",
     main=paste("Histogram of Patient's Age vs Frequency "),col="green",
     axes = TRUE, 
     labels = TRUE, ylim=c(0,400),
     xlim = c(0, 18),xaxp=c(0,17,17),
     yaxp=c(0,400,8))
```
```{r}
aggregate(TOTCHG ~ AGE, FUN = sum, data = hos_cost)
```

$\color{blue}{\text {THIS HISTOGRAM SHOWS 0 AGE PEOPLE VISITED MOST FREQUENTLY.}}$

$\color{blue}{\text {Expenditure: 0 age people is 678118 and 15 years age cost 111747.}}$

$\color{blue}{\text {Expenditure wise cost:0 age people 678118	and 17 years age people cost 174777}}$
\newpage

## 2. IN ORDER OF SEVERITY OF THE DIAGNOSIS AND TREATMENTS AND TO FIND OUT THE EXPENSIVE TREATMENTS, THE AGENCY WANTS TO FIND THE DIAGNOSIS-RELATED GROUP THAT HAS MAXIMUM HOSPITALIZATION AND EXPENDITURE.
```{r}
#TO FIND THE EXPENDITURE OF by category
hist(hos_cost$APRDRG,freq = TRUE,right = TRUE,
     border = "blue",
  main=paste("Histogram of APRDRG vs Frequency"), col="pink",
  axes = TRUE, xlab="APRDRG",
  ylab = "Frequency", ylim =c(0,300),
     xlim = c(0, 1000),xaxp=c(0,1000,10),
     yaxp=c(0,300,6))
```
\newpage

$\color{blue}{\text {In category 600-700 people have visited hospital frequently.}}$

# let's find which number related dignosis group visited that.
```{r}
factor_cost <-as.factor(hos_cost$APRDRG)
summary(factor_cost)
which.max(summary(factor_cost))
```
# THIS SHOW 640 DIAGNISTIC GROUP HAS MAXIMUM EXPENDITURE AND MAXIMUM HOSPITALIZATION
\newpage
## 3. TO MAKE SURE THAT THERE IS NO MALPRACTICE, THE AGENCY NEEDS TO ANALYZE IF THE RACE OF THE PATIENT IS RELATED TO THE HOSPITALIZATION COSTS.
We are interested in anova analysis.
H0:All means are equal among races. There is no malpractice.
H1:There is some biasness and people are treated according with race.
Known fact:if p value<0.05 : H0 is rejected, H1 is accepted
P value>0.05 , H0 is accepted
Degrees of freedom: df(num)=k-1 # k= number of groups has been tested here 
```{r}
hos_cost2<-na.omit(hos_cost)
factor_race <-as.factor(hos_cost2$RACE)
summary(factor_race)
# KEY IDEAS
# 6 RACES SO 5 DEGREES OF FREEDOM
#Calculate the test statistic: F=2.23
#Define probability statement: p-value=P(F>2.23)=0.1241
#Decide: IF ALPHA < P, ACCEPT H0
ano_model <- aov(TOTCHG~factor_race,data=hos_cost2)
summary(ano_model)
```
$\color {blue} {\text {p value (0.943) > 0.05 so H0 is accepted. The Residual Value (deviation of the observed value) is very }}$
$\color {blue} {\text {high specifying that there is no relation between the race of patient and the hospital cost.}}$ $\color {blue} {\text {From the summary we can also see that the data has 484 patients of Race 1 out of the 500 entries.}}$ 
$\color {blue} {\text {This will affect the results of ANOVA as well, since the number of observations is very much skewed.}}$ 
$\color {blue} {\text { Hence we can conclude that there is no race wise cost bias in the observed data.}}$ 
\newpage

## 4.TO PROPERLY UTILIZE THE COSTS, THE AGENCY HAS TO ANALYZE THE SEVERITY OF THE HOSPITAL COSTS BY AGE AND GENDER FOR THE PROPER ALLOCATION OF RESOURCES.
# To analyse whether gender and age are related with hospital cost. Linear regression model is useful.
```{r}
lm_model <- lm(hos_cost$TOTCHG ~ hos_cost$AGE+hos_cost$FEMALE)
summary(lm_model)
```
\newpage
$\color{blue}{\text{CONCLUSION}}$ 

# P VALUES OF MODEL are LESS THAN 0.05 ARE AGE,gender &CORRESPONDING SLOPES ARE 86.28 ,-748.19
# AGE is POSITIVELY RELATED. So older patient cost is more.
# Female IS NEGATIVELY RELATED.so being female pay less cost than male.
\newpage

## 5. SINCE THE LENGTH OF STAY IS THE CRUCIAL FACTOR FOR INPATIENTS, THE AGENCY WANTS TO FIND IF THE LENGTH OF STAY CAN BE PREDICTED FROM AGE, GENDER, AND RACE.
# To analyse whether length of stay is related with age,gender and race. multiple Linear regression model is useful.

```{r}
lm_model2 <- lm(hos_cost$LOS ~ hos_cost$AGE+hos_cost$FEMALE +hos_cost$RACE)
summary(lm_model2)
```
$\color{blue}{\text{CONCLUSION}}$ 

# The significance codes are almost null for all the variables, except for the intercept. 

# The p-value high which signifies that there is no linear relationship between the given variables. 
# Hence we cannot predict the length of stay of the patients based on the age, gender, and race.

\newpage
## 6. TO PERFORM A COMPLETE ANALYSIS, THE AGENCY WANTS TO FIND THE VARIABLE THAT MAINLY AFFECTS HOSPITAL COSTS.
# To analyse whether hospital is related with all other variables. multiple Linear regression model is useful.
```{r}
model2<- lm(TOTCHG ~.,data = hos_cost)
summary(model2)

```
$\color{blue}{\text{CONCLUSION}}$ 

# Looking p values hospital cost is mostly depends on age (7.02e-14 < 0.05) and slope is 134.6949. This implies older age people have higher hospital cost.

# p value of LOS is also affecting factor. as P VALUE 2e-16<0.05. SLOPE IS HIGHER POSITIVE 743.1521. iMPLIES HIGHER LENGTH OF STAY COSTS MORE. 

# Another important factor affecting cost is All Patient Refined Diagnosis Related Groups. p value is 2e-16 < 0.05 and slope is -7.7909. This group affecting in negative way.  










